---
title: "seedbank_mutation_check"
author: "Joy O'Brien"
date: "2026-04-06"
output: html_document
---
Goal: Determine if there are any mutations that arise in phage that coevolved with a non-sporulating host compared to a sporulating host; if entrapment is under strong genetic control then passaging on non sporulatio should relax selection on auxiliary sporulation elements
```{r}
# setwd
setwd("~/GitHub/VirosporeQuant/seedbank.check")
```


0. Set theme
```{r}
library(paletteer)
library(ggplot2)
theme_set(theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.title.x = element_text(margin = margin(t = 15, b = 15)),
        axis.title.y = element_text(margin = margin(l = 15, r = 15)),
        axis.text = element_text(size = 13),
        axis.text.x = element_text(margin = margin(t = 5)),
        axis.text.y = element_text(margin = margin(r = 5)),
        #axis.line.x = element_line(linewidth = 1),
        #axis.line.y = element_line(linewidth = 1),
        axis.ticks.y.right = element_line(linewidth = 1),
        axis.text.y.right = element_text(size = 13, margin = margin(l = 5)),
        axis.line.y.right = element_line(linewidth = 0.1),
        axis.ticks.x.top = element_line(linewidth = 1),
        axis.text.x.top = element_text(size = 13, margin = margin(b = 5)),
        axis.line.x.top = element_line(linewidth = 0.1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_line(linewidth = 1),
        axis.ticks.y = element_line(linewidth = 1),
        axis.ticks.length = unit(.1, "in"),
        panel.border = element_rect(color = "grey20", fill = NA, linewidth = 1.5),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14),
        strip.background = element_blank()
        )) 
colors <-paletteer_d("ggthemes::Miller_Stone")
```

1. Load in data
```{r}
WLO.l1 <- read.csv("~/GitHub/VirosporeQuant/seedbank.check/Data/WLO_L1_phage_long_seed_bank_SPO1_filtered_phage_annotated_timecourse.txt", header = TRUE)
WLO.l2 <- read.csv("~/GitHub/VirosporeQuant/seedbank.check/Data/WLO_L2_phage_long_seed_bank_SPO1_filtered_phage_annotated_timecourse.txt", header = TRUE)
WLO.l3 <- read.csv("~/GitHub/VirosporeQuant/seedbank.check/Data/WLO_L3_phage_long_seed_bank_SPO1_filtered_phage_annotated_timecourse.txt", header = TRUE)

SNO.l1 <- read.csv("~/GitHub/VirosporeQuant/seedbank.check/Data/SNO_L1_phage_no_seed_bank_SPO1_filtered_phage_annotated_timecourse.txt", header = TRUE)
SNO.l2 <- read.csv("~/GitHub/VirosporeQuant/seedbank.check/Data/SNO_L2_phage_no_seed_bank_SPO1_filtered_phage_annotated_timecourse.txt", header = TRUE)
SNO.l3 <- read.csv("~/GitHub/VirosporeQuant/seedbank.check/Data/SNO_L3_phage_no_seed_bank_SPO1_filtered_phage_annotated_timecourse.txt", header = TRUE)
```


2. Clean up and wrangle data
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

WLO.l1$Dataset <- "Spore.lysate1"
WLO.l2$Dataset <- "Spore.lysate2"
WLO.l3$Dataset <- "Spore.lysate3"
SNO.l1$Dataset <- "Nonspore.lysate1"
SNO.l2$Dataset <- "Nonspore.lysate2"
SNO.l3$Dataset <- "Nonspore.lysate3"

spore.mut <- bind_rows(WLO.l1, WLO.l2, WLO.l3)
nonspore.mut <- bind_rows(SNO.l1, SNO.l2, SNO.l3)

# Add labels for treatment
spore.mut$Coevolve <- "Spore"
nonspore.mut$Coevolve <- "Non-spore"

all.mut <- bind_rows(spore.mut, nonspore.mut)

# Reshape frequencies
long.mut <- all.mut %>%
  pivot_longer(
    cols = starts_with("Freq."),
    names_to = "Timepoint",
    values_to = "Frequency"
  )
# Align mutations across treatments
```


```{r}
#all_mut <- bind_rows(WLO.l1, WLO.l2, WLO.l3, SNO.l1, SNO.l2, SNO.l3)

#all_mut_long <- all_mut %>%
  # pivot_longer(
  #   cols = starts_with("Freq."),
  #   names_to = "Timepoint",
  #   values_to = "Frequency"
  # )
```

```{r}
ggplot(all_mut_long, aes(x = Position, y = Frequency, color = Dataset)) +
  geom_point(alpha = 0.6, size = 1.5) +
  facet_wrap(~Timepoint, ncol = 1) +
  theme_bw()
```


2. Plot
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)


mut_summary <- long.mut %>%
  group_by(Gene, Coevolve, Timepoint) %>%
  summarize(
    mean_freq = mean(Frequency, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(long.mut, aes(x = Gene, y = Frequency, fill = Coevolve)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Gene",
    y = "Mutation Frequency",
    fill = "Treatment"
  )

# Edit the order of frequency in the plot 
mut_summary <- mut_summary %>%
  mutate(
    Timepoint = factor(Timepoint,
                       levels = c("Freq.1","Freq.4","Freq.7","Freq.10","Freq.14"))
  )
# Gene locus that contain our elements of interest --> 0A and par-like elements
ofinterest <- data.frame(
  Gene = c("SPO1_55", "SPO1_136", "SPO1_97", "SPO1_141", "SPO1_160", "SPO101"))
  #Element = c("ParB", "ParM", "0A_box", "0A_box", "0A_box", "0A_box")
#)
  
ggplot(mut_summary, aes(x = Timepoint, y = Gene, fill = mean_freq)) +
  geom_tile() +
  facet_wrap(~Coevolve) +
  scale_fill_viridis_c() + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
# Just realized the genes are actually locus tags
```

3. Plot specifically for elements we are interested in
```{r}
# Gene locus that contain our elements of interest --> 0A and par-like elements
ofinterest <- data.frame(
  Gene = c("SPO1_55", "SPO1_136", "SPO1_97", "SPO1_141", "SPO1_160", "SPO101"))
  #Element = c("ParB", "ParM", "0A_box", "0A_box", "0A_box", "0A_box")
#)

# your long plotting dataframe
# example columns: Gene, Timepoint, Treatment, Frequency
mut_summary_int<- mut_summary_int %>%
  mutate(is_ofinterest = Gene %in% ofinterest$Gene) %>%
  mutate(
    Gene_label = ifelse(
      is_ofinterest,
      paste0("<span style='color:red;'>", Gene, "</span>"),
      Gene
    )
  )
          

ggplot(mut_summary_int, aes(x = Timepoint, y = Gene_label, fill = mean_freq)) +
  geom_tile() +
  facet_wrap(~Coevolve) +
  scale_fill_viridis_c() + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(mut_summary_int, aes(x = Timepoint, y = Gene_label, fill = mean_freq)) +
  geom_tile() 
```
4. Determine the location of the mutations within genes that carry 0A boxes
```{r}


```

