---
title: "SPO1 wt optimal entrapment analysis"
author: "Joy O'Brien"
date: "2026-04-24"
output: html_document
---
1. Set wd and figure theme
```{r}
setwd("~/GitHub/VirosporeQuant/Optimal.entrap")
```

```{r}
library(paletteer)
library(ggplot2)
theme_set(theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.title.x = element_text(margin = margin(t = 15, b = 15)),
        axis.title.y = element_text(margin = margin(l = 15, r = 15)),
        axis.text = element_text(size = 13),
        axis.text.x = element_text(margin = margin(t = 5)),
        axis.text.y = element_text(margin = margin(r = 5)),
        #axis.line.x = element_line(linewidth = 1),
        #axis.line.y = element_line(linewidth = 1),
        axis.ticks.y.right = element_line(linewidth = 1),
        axis.text.y.right = element_text(size = 13, margin = margin(l = 5)),
        axis.line.y.right = element_line(linewidth = 0.1),
        axis.ticks.x.top = element_line(linewidth = 1),
        axis.text.x.top = element_text(size = 13, margin = margin(b = 5)),
        axis.line.x.top = element_line(linewidth = 0.1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_line(linewidth = 1),
        axis.ticks.y = element_line(linewidth = 1),
        axis.ticks.length = unit(.1, "in"),
        panel.border = element_rect(color = "grey20", fill = NA, linewidth = 1.5),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14),
        strip.background = element_blank()
        )) 
colors <-paletteer_d("ggthemes::Miller_Stone")
```

2. Load cell synchrony data --> note to self need to divide cell counts by 7 to account for the concentrating of the sample in the lab (7 mL --> 1 mL)
```{r}
library(readxl)
sync <- read_excel("~/GitHub/VirosporeQuant/Optimal.entrap/20260626_1522_cellsynchrony_github.xlsx", sheet = "sync")
```

2. Calculate spore fraction of cells and plot to get initial sense of data
```{r}
library(tidyverse)
sync<- sync %>%
  mutate(
    spore_fraction = spores.ml / (spores.ml + veg.cells.ml)
  )

ggplot(sync, aes(x = `Time (h)`, y = spore_fraction, group = Replicate)) +
  geom_line(aes(color = as.factor(Replicate)), size = 1) +
  #geom_point(size = 2) +
  labs(
    x = "Time (h)",
    y = "Spore fraction (spores / total cells)",
    color = "Replicate"
  )
```

3. Format long and plot summary
```{r}
sync_long <- sync %>%
  select(`Time (h)`, Replicate, veg.cells.ml, spores.ml, OD600) %>%
  pivot_longer(
    cols = c(veg.cells.ml, spores.ml, OD600),
    names_to = "variable",
    values_to = "value"
  )
ggplot(sync_long, aes(x = `Time (h)`, y = value, color = variable)) +
  geom_line(aes(group = interaction(variable, Replicate)), alpha = 0.6) +
  geom_point(alpha = 0.6) +
  scale_y_log10() +
  labs(
    x = "Time (hours)",
    y = "Value (log scale)",
    color = "Measurement"
  )
```

4. Calculate the mean spore fraction 
```{r}
sync_summary <- sync %>%
  group_by(`Time (h)`) %>%
  summarize(
    mean_spore = mean(spore_fraction),
    se_spore = sd(spore_fraction)/sqrt(n())
  )
ggplot(sync_summary, aes(x = `Time (h)`, y = mean_spore)) +
  geom_line(size = 1.2, color = "#d95f02") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_spore - se_spore,
                    ymax = mean_spore + se_spore),
                width = 0.2) +
  labs(
    x = "Time (h)",
    y = "Mean spore fraction"
  ) 
```
5. Plot growth curve part of the data (OD600)
```{r}
library(tidyverse)
ggplot(sync, aes(x = `Time (h)`, y = OD600)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  labs(
    x = "Time (h)",
    y = "OD600"
  ) 
```

6. Plot vegetative cells and spores together
```{r}
# Convert to long format
sync_counts_long <- sync %>%
  select(`Time (h)`, Replicate, veg.cells.ml, spores.ml) %>%
  pivot_longer(
    cols = c(veg.cells.ml, spores.ml),
    names_to = "Cell_type",
    values_to = "cells_per_ml"
  ) %>%
  mutate(
    Cell_type = recode(
      Cell_type,
      veg.cells.ml = "Vegetative cells",
      spores.ml = "Spores"
    )
  )
sync_counts_long <- sync_counts_long %>% # reorder vegetative cells and spores so that they appear correctly in the figure
  mutate(
    Cell_type = factor(Cell_type,
                       levels = c("Vegetative cells", "Spores"))
  )
# Plot 
ggplot(sync_counts_long, aes(x = `Time (h)`, y = cells_per_ml, color = Cell_type)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Cells or spores per mL",
    color = ""
  )

# Replace 0 with NA 
sync_counts_long <- sync_counts_long %>%
  mutate(cells_per_ml_plot = if_else(cells_per_ml > 0, cells_per_ml, NA_real_))

ggplot(sync_counts_long, aes(x = `Time (h)`, y = cells_per_ml_plot, color = Cell_type)) +
  stat_summary(fun = mean, geom = "line", linewidth = 1.2, na.rm = TRUE) +
  stat_summary(fun = mean, geom = "point", size = 3, na.rm = TRUE) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, na.rm = TRUE) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Cells or spores per mL",
    color = ""
  )
```

7. Follow up stats
```{r}
# Spore counts over time
kruskal.test(spores.ml ~ factor(`Time (h)`), data = sync)

# Spore fraction over time
kruskal.test(spore_fraction ~ factor(`Time (h)`), data = sync)

library(FSA)

dunnTest(
  spores.ml ~ factor(`Time (h)`),
  data = sync,
  method = "holm"
)

dunnTest(
  spore_fraction ~ factor(`Time (h)`),
  data = sync,
  method = "holm"
)

# veg cells
kruskal.test(
  corr.veg.cells.ml ~ factor(`Time (h)`),
  data = sync
)
```

8. Determining time of max sporulation rate --does spore fraction significantly change over time? 
```{r}
sync_summary <- sync %>%
  group_by(`Time (h)`) %>%
  summarize(
    mean_spore_fraction = mean(spores.ml / (spores.ml + veg.cells.ml), na.rm = TRUE),
    se = sd(spores.ml / (spores.ml + veg.cells.ml), na.rm = TRUE) / sqrt(n())
  ) %>%
  arrange(`Time (h)`) %>%
  mutate(
    delta_spore_fraction = mean_spore_fraction - lag(mean_spore_fraction)
  )

sync <- sync %>%
  mutate(spore_fraction = spores.ml / (spores.ml + veg.cells.ml))

mod <- aov(spore_fraction ~ factor(`Time (h)`), data = sync)
summary(mod)


```

9. Initial/pilot optimal entrapment data
```{r}
library(readxl)
opt <- read_excel("~/GitHub/VirosporeQuant/Optimal.entrap/20260629_optimalentrapment_experiment_data_github.xlsx", sheet = "opt.e")
```

As of 6/30-plot spores over time for each moi (only have spore data)
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
# Convert to long format
opt_long <- opt %>%
  select(Infection.time.h, Rep, spores.ml, virospores.ml) %>%
  pivot_longer(
    cols = c(spores.ml, virospores.ml),
    names_to = "Type",
    values_to = "count"
  ) %>%
  mutate(
    Type = recode(Type,
                  spores.ml = "Spores",
                  virospores.ml = "Virospores")
  )
# Reorder virospores and spores so it makes sense on the plot
opt_long <- opt_long %>%
  mutate(
    Type = factor(Type,
                       levels = c("Virospores", "Spores"))
  )
# First look
ggplot(opt, aes(x =Infection.time.h, y = spores.ml)) +
  geom_line(aes(group = Rep, color = as.factor(Rep)), size = 1) +
  geom_point(size = 2) +
  labs(
    x = "Time of infection (hours)",
    y = "Spores per mL",
    color = "Replicate"
  )

ggplot(opt_long, aes(x = Infection.time.h, y = count, color = Type)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Cells per mL",
    color = ""
  )
```

10. Calculate summary stats
```{r}
# Calcaulte summary stats
spore_summary <- opt %>%
  group_by(Infection.time.h, MOI) %>%
  summarise(
    mean_spores = mean(spores.ml, na.rm = TRUE),
    sd_spores = sd(spores.ml, na.rm = TRUE),
    n = sum(!is.na(spores.ml)),
    se_spores = sd_spores / sqrt(n),
    .groups = "drop"
  )

# Plot 
ggplot(spore_summary,
       aes(x = Infection.time.h,
           y = mean_spores,
           color = factor(MOI),
           group = factor(MOI))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = mean_spores - se_spores,
        ymax = mean_spores + se_spores),
    width = 0.3
  ) +
  #scale_y_log10() +        # been playing around with removing or not 
  labs(
    x = "Time of Infection (h)",
    y = "Spores per mL",
    color = "MOI"
  )
```


11.Instead of calculating mean spores and se_spores on raw scale, each replicate can be converted log10 (scaling the data) THEN calculate mean and SE on the log scale, then convert back to spores per ml -- i.e., transformed each replicate to log10 averaged the transformed values, calculated error on transformed scale, then converted back to spores/mL for plotting. 
```{r}
spore_summary_log <- opt %>%
  filter(spores.ml > 0) %>% # remove 0
  mutate(log_spores = log10(spores.ml)) %>% #transform the replicate
  group_by(Infection.time.h, MOI) %>% #group by 
  summarise(
    mean_log = mean(log_spores, na.rm = TRUE), #calculate average on log scale
    se_log = sd(log_spores, na.rm = TRUE) / sqrt(n()), #calculate se
    ymin = 10^(mean_log - se_log),
    ymax = 10^(mean_log + se_log),
    mean_spores = 10^mean_log,
    .groups = "drop"
  )

panel_A <- ggplot(spore_summary_log,
       aes(x = Infection.time.h,
           y = mean_spores,
           color = factor(MOI),
           group = factor(MOI))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = ymin, ymax = ymax),
    width = 0.3
  ) +
  scale_y_log10() +
  labs(
    x = "Time of Infection (h)",
    y = "Spores per mL",
    color = "MOI"
  ) 
panel_A
```

Checking data distribution 
```{r}
opt$log_spores <- log10(opt$spores.ml)
log_data <- opt %>%
  filter(spores.ml > 0) %>%
  mutate(log_spores = log10(spores.ml))

qqnorm(log_data$log_spores)
qqline(log_data$log_spores, col = "red")
```

12. Preliminary stats on uninfected spores
```{r}
# in order to do anova you need to have the following: 
# indepdendent obs 
# normality 
# homogeneity of variance
# continuous dependent variable

# First determine if ANOVA is appropriate
model <- lm(log10(spores.ml + 1) ~ factor(Infection.time.h) * factor(MOI),
            data = opt)
anova(model)
par(mfrow = c(2,2))
plot(model)

# test the normality of residuals
shapiro.test(residuals(model))
```

Confirm variables and missing data
```{r}
str(opt)

colSums(is.na(opt[, c(
  "spores.ml",
  "Infection.time.h",
  "MOI"
)]))

# Remove virospore columns for now 
library(dplyr)
#opt <- opt %>%
  #select(-vPFU.count, -vPFU.dilution)

# visualize raw distributions
library(ggplot2)

ggplot(
  opt,
  aes(
    x = interaction(Infection.time.h, MOI),
    y = spores.ml
  )
) +
  geom_boxplot() +
  geom_jitter(width = 0.15) +
  scale_y_log10() +
  labs(
    x = "Infection time × MOI",
    y = "Spores per mL"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

Fit anova model
```{r}
opt <- opt %>%
  mutate(
    Infection.time.h = factor(Infection.time.h),
    MOI = factor(MOI)
  )

model <- lm(
  log10(spores.ml + 1) ~ Infection.time.h * MOI,
  data = opt
)

# Check residuals
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))

# Check residual normality
qqnorm(residuals(model))
qqline(residuals(model))

shapiro.test(residuals(model)) # residuals are not normal

# Check equal variances
library(car)

leveneTest(
  log10(spores.ml + 1) ~
    interaction(Infection.time.h, MOI),
  data = opt
)
# Okay variances are equal


#Visually check the residual spread 
plot(
  fitted(model),
  residuals(model),
  xlab = "Fitted values",
  ylab = "Residuals"
)

abline(h = 0, lty = 2)

# Check for unusual obs
cooks <- cooks.distance(model)

which(cooks > 4 / nobs(model))
```

Moving forward with ANOVA
```{r}
# Fit the full model
model1 <- lm(
  log10(spores.ml + 1) ~ factor(Infection.time.h) * factor(MOI),
  data = opt
)

anova(model)
# okay so timing of infection is important and so is MOI but because the interaction between time and MOI is not significant, we can transition to an additive model 

# We can simplify the model because time and MOI were not signififcant 
model2 <- lm(
  log10(spores.ml + 1) ~ Infection.time.h + MOI, 
  data = opt
)

anova(model2)
# Post hoc tests
library(emmeans)
emm_time <- emmeans(model2, ~ Infection.time.h)
pairs(emm_time, adjust = "tukey")

# compact letter display
library(multcomp)

cld(
  emm_time,
  adjust = "tukey",
  Letters = letters
)
# Okay need to add a panel to the figure of the distribtuons of MOI-- the marginal means from emmeans for infection time time averaged across MOIOS with the compact letter display 

library(emmeans)
library(multcomp)
library(dplyr)
library(ggplot2)

# Add Tukey compact-letter groups
emm_time_letters <- cld(
  emm_time,
  adjust = "tukey",
  Letters = letters
) %>%
  as.data.frame() %>%
  mutate(
    Infection.time.h = as.numeric(as.character(Infection.time.h)),
    .group = trimws(.group)
  )

panel_B <- ggplot(
  emm_time_letters,
  aes(x = Infection.time.h, y = emmean)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.3
  ) +
  geom_text(
    aes(y = upper.CL + 0.25, label = .group),
    size = 4
  ) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 2)
  ) +
  labs(
    x = "Infection time (h)",
    y = expression("Estimated marginal mean log"[10] *
                     "(spores mL"^{-1} * " + 1)")
  ) 

panel_B

# Because the model works on the log scale, we need to back track to spores/ML
emm_time_letters_original <- emm_time_letters %>%
  mutate(
    spores_emmean = 10^emmean - 1,
    spores_lower = 10^lower.CL - 1,
    spores_upper = 10^upper.CL - 1
  )

# Plot on tthe original scale 
panel_B <- ggplot(
  emm_time_letters_original,
  aes(x = Infection.time.h, y = spores_emmean)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = spores_lower, ymax = spores_upper),
    width = 0.3
  ) +
  geom_text(
    aes(y = spores_upper * 1.4, label = .group),
    size = 4
  ) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 2)
  ) +
  scale_y_log10() +
  labs(
    x = "Infection time (h)",
    y = "Model-adjusted mean spores per mL"
  ) 

panel_B
# MOI was also significant
emm_moi <- emmeans(model, ~ MOI)
pairs(emm_moi, adjust = "tukey")

moi_results <- cld(
  emm_moi,
  adjust = "tukey",
  Letters = letters
) %>%
  as.data.frame() %>%
  mutate(
    .group = trimws(.group),
    estimated_spores = 10^emmean - 1,
    lower_spores = 10^lower.CL - 1,
    upper_spores = 10^upper.CL - 1
  )
# Plot MOIS
panel_C <- ggplot(
  moi_results,
  aes(x = MOI, y = estimated_spores)
) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower_spores, ymax = upper_spores),
    width = 0.15
  ) +
  geom_text(
    aes(y = upper_spores * 1.3, label = .group),
    size = 4
  ) +
  scale_y_log10() +
  labs(
    x = "MOI",
    y = "Model-adjusted mean spores per mL"
  ) 

panel_C

library(patchwork)

final_figure <- panel_A / (panel_B + panel_C) +
  plot_annotation(tag_levels = "A")

final_figure

ggsave(
  "Figure1_spores.pdf",
  width = 10,
  height = 8,
  dpi = 300
)

```



13. Putting the panel figures together
```{r}
library(patchwork)

combined_figure <- panel_A + panel_B +
  plot_annotation(tag_levels = "A")

combined_figure
```
14. Modeling time as continuous
```{r}


model3 <- lm(log10(spores.ml + 1) ~ Infection.time.h, data = opt)
model3
anova(model3)

model4 <- lm(log10(spores.ml + 1) ~ Infection.time.h + MOI, data = opt)
anova(model4)
```

```{r}
install.packages("drc")
library(drc)

# verify that MOI is a factor
log_data$MOI <- as.factor(log_data$MOI) #using the log transformed data because counts spanned more than 1 order of magnitude 

# Okay tying to fit 4-parameter log-logistic /sigmoidal curve per moi using drc
model_full <- drm(log_spores ~ Infection.time.h, curveid = MOI, data = log_data, fct = L.4())

summary(model_full)
plot(model_full, log = "", legendPos = c(2, max(log_data$log_spores)))

# Test whether curves differ between MOIs at all 
# fit reduced model where all MOIs share the same curve 
model_reduced <- drm(log_spores ~ Infection.time.h, data = log_data, fct = L.4())
summary(model_reduced)

plot(model_reduced, log = "", legendPos = c(2, max(log_data$log_spores)))

anova(model_reduced, model_full)
compParm(model_full, "e", "-")   # pairwise differences in Xmid between MOI groups
compParm(model_full, "b", "-")   # pairwise differences in slope
compParm(model_full, "d", "-")   # pairwise differences in top/plateau


model_e_only <- drm(slogpores ~ time, curveid = moi, data = df, fct = L.4(),
                     pmodels = list(~1, ~1, ~1, ~moi))  # b,c,d shared; e (Xmid) free per MOI
anova(model_e_only, model_full)

```

15. Calculate entrapment efficiency, format long 
```{r}
# EE 
opt <- opt %>%
  mutate(
    ee = (virospores.ml / (virospores.ml + spores.ml)) * 100
  )

opt_long <- opt %>%
  select(Infection.time.h, Rep, spores.ml, virospores.ml, ee) %>%
  pivot_longer(
    cols = c(spores.ml, virospores.ml),
    names_to = "Type",
    values_to = "count"
  ) %>%
  mutate(
    Type = recode(Type,
                  spores.ml = "Spores",
                  virospores.ml = "Virospores")
  )
# reorder appearance of virospores and spores
opt_long <- opt_long %>%
  mutate(
    Type = factor(Type,
                       levels = c("Virospores", "Spores"))
  )
  
```
Plot entrapment
```{r}
ggplot(opt_long, aes(x = Infection.time.h, y = count, color = Type)) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Cells per mL",
    color = ""
  ) 
```
Plot ee over time 
```{r}
ggplot(opt_long, aes(x = Infection.time.h, y = ee)) +
  geom_line(aes(group = Rep, color = as.factor(Rep)), size = 1) +
  geom_point(size = 2) +
  labs(
    x = "Time of infection (hours)",
    y = "Entrapment efficiency (%)",
    color = "Replicate"
  )
```

```{r}
ggplot(opt_long, aes(x = Infection.time.h, y = ee)) +
  stat_summary(fun = mean, geom = "line", size = 1.2, color = "#7570b3") +
  stat_summary(fun = mean, geom = "point", size = 3, color = "#7570b3") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "#7570b3") +
  labs(
    x = "Time of infection (h)",
    y = "Entrapment efficiency (%)"
  ) 
```

PART TWO: WORKING WITH UPDATED DATA (other analyses are in the optimal.entrap.tempr.Rmd because I had to work off posit offline when I didn't have my laptop)
```{r}
library(readxl)
opt <- read_excel("~/GitHub/VirosporeQuant/Optimal.entrap/Data/20260526_2205_optentrap_expdesign.xlsx", sheet = "CFU_vPFU")
```

Looking at distribution of virospores via boxplot (not the right way to plot)
```{r}
library(tidyverse)

ggplot(opt, aes(x = factor(Time.infection.hr), y = virospores.ml)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +
  geom_jitter(aes(color = factor(MOI)), width = 0.15, size = 2, alpha = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Virospores per mL",
    color = "MOI",
    title = "Distribution of virospores over time"
  )
```
Summarize data and plot 
```{r}
opt_summary <- opt %>%
  group_by(Time.infection.hr, MOI) %>%
  summarise(
    mean_virospores = mean(virospores.ml, na.rm = TRUE),
    se_virospores = sd(virospores.ml, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(opt_summary, aes(x = Time.infection.hr, y = mean_virospores, color = factor(MOI))) +
  geom_line() +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = mean_virospores - se_virospores,
        ymax = mean_virospores + se_virospores),
    width = 0.2
  ) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Virospores per mL",
    color = "MOI"
  ) 
```
Exploring different ways to plot
```{r}
ggplot(opt,
       aes(x = factor(Time.infection.hr),
           y = virospores.ml,
           fill = factor(MOI))) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1,
               fill = "white") +
  scale_y_log10()
```
```{r}
ggplot(opt,
       aes(x = factor(Time.infection.hr),
           y = virospores.ml,
           fill = factor(Rep))) +
  geom_col(position = position_dodge(width = 0.8)) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Virospores per mL",
    fill = "Replicate"
  )
```

```{r}

opt_summary3 <- opt %>%
  group_by(Time.infection.hr) %>%
  summarise(
    mean_virospores = mean(virospores.ml),
    se = sd(virospores.ml)/sqrt(n()),
    .groups = "drop"
  )
ggplot(opt_summary3,
       aes(x = factor(Time.infection.hr),
           y = mean_virospores)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = mean_virospores - se,
        ymax = mean_virospores + se),
    width = 0.2
  ) +
  geom_point(data = opt,
             aes(x = factor(Time.infection.hr),
                 y = virospores.ml),
             inherit.aes = FALSE,
             position = position_jitter(width = 0.1),
             size = 2) +
  scale_y_log10()
```

