---
title: "SPO1 wt optimal entrapment analysis"
author: "Joy O'Brien"
date: "2026-04-24"
output: html_document
---
1. Set wd and figure theme
```{r}
setwd("~/GitHub/VirosporeQuant/Optimal.entrap")
```

```{r}
library(paletteer)
library(ggplot2)
theme_set(theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.title.x = element_text(margin = margin(t = 15, b = 15)),
        axis.title.y = element_text(margin = margin(l = 15, r = 15)),
        axis.text = element_text(size = 13),
        axis.text.x = element_text(margin = margin(t = 5)),
        axis.text.y = element_text(margin = margin(r = 5)),
        #axis.line.x = element_line(linewidth = 1),
        #axis.line.y = element_line(linewidth = 1),
        axis.ticks.y.right = element_line(linewidth = 1),
        axis.text.y.right = element_text(size = 13, margin = margin(l = 5)),
        axis.line.y.right = element_line(linewidth = 0.1),
        axis.ticks.x.top = element_line(linewidth = 1),
        axis.text.x.top = element_text(size = 13, margin = margin(b = 5)),
        axis.line.x.top = element_line(linewidth = 0.1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_line(linewidth = 1),
        axis.ticks.y = element_line(linewidth = 1),
        axis.ticks.length = unit(.1, "in"),
        panel.border = element_rect(color = "grey20", fill = NA, linewidth = 1.5),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14),
        strip.background = element_blank()
        )) 
colors <-paletteer_d("ggthemes::Miller_Stone")
```


1. Optimal entrapment analysis for SPO1 WT phage 
```{r}
library(readxl)
opt <- read_excel("~/GitHub/VirosporeQuant/Optimal.entrap/Data/20260629_optimalentrapment_experiment_data_github.xlsx", sheet = "opt.e")
```

2. Calculate total spores and entrapment efficiency
```{r}
library(tidyverse)
library(scales)

opt.df <- opt %>%
  mutate(
    MOI = factor(MOI),
    total_spores_ml = virospores.ml.corrected + spores.ml.corrected,
    entrapment_efficiency = if_else(
      !is.na(total_spores_ml) & total_spores_ml > 0,
      virospores.ml.corrected / total_spores_ml,
      NA_real_
    ),
    entrapment_percent = 100 * entrapment_efficiency
  )
```

3. Summarize data (entrapment efficiency) and plot
```{r}
efficiency_summary <- opt.df %>%
  group_by(MOI, Infection.time.h) %>%
  summarise(
    n = sum(!is.na(entrapment_percent)),
    mean_EE = mean(entrapment_percent, na.rm = TRUE),
    sd_EE = sd(entrapment_percent, na.rm = TRUE),
    se_EE = sd_EE / sqrt(n),
    .groups = "drop"
  )

ggplot() +
  geom_point(
    data = opt.df,
    aes(
      x = Infection.time.h,
      y = entrapment_percent,
      color = MOI
    ),
    alpha = 0.45,
    position = position_jitter(width = 0.08)
  ) +
  geom_line(
    data = efficiency_summary,
    aes(
      x = Infection.time.h,
      y = mean_EE,
      color = MOI,
      group = MOI
    ),
    linewidth = 0.9
  ) +
  geom_point(
    data = efficiency_summary,
    aes(
      x = Infection.time.h,
      y = mean_EE,
      color = MOI
    ),
    size = 2.5
  ) +
  geom_errorbar(
    data = efficiency_summary,
    aes(
      x = Infection.time.h,
      ymin = pmax(0, mean_EE - se_EE),
      ymax = pmin(100, mean_EE + se_EE),
      color = MOI
    ),
    width = 0.15
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    labels = label_number(suffix = "%")
  ) +
  labs(
    x = "Infection time (h)",
    y = "Entrapment efficiency (%)",
    color = "MOI",
    title = "Entrapment efficiency by infection time and MOI"
  ) 
```

```{r}
# Investigating missing sample; checking for problematic entrapment values
opt.df %>%
  filter(
    !is.finite(entrapment_percent) |
      entrapment_percent < 0 |
      entrapment_percent > 100
  ) %>%
  select(
    Sample, Infection.time.h, Rep, MOI,
    spores.ml, virospores.ml, total_spores_ml,
    entrapment_percent
  )
# Create clean dataset
plot_results <- opt.df %>%
  filter(
    is.finite(Infection.time.h),
    is.finite(entrapment_percent),
    between(entrapment_percent, 0, 100)
  )

# Calculate summary statistics
efficiency_summary <- plot_results %>%
  group_by(MOI, Infection.time.h) %>%
  summarise(
    n = n(),
    mean_EE = mean(entrapment_percent),
    sd_EE = if_else(n > 1, sd(entrapment_percent), NA_real_),
    se_EE = sd_EE / sqrt(n),
    .groups = "drop"
  )
# Plot
ggplot() +
  geom_point(
    data = plot_results,
    aes(
      x = Infection.time.h,
      y = entrapment_percent,
      color = MOI
    ),
    alpha = 0.45,
    position = position_jitter(width = 0.08, height = 0)
  ) +
  geom_line(
    data = efficiency_summary,
    aes(
      x = Infection.time.h,
      y = mean_EE,
      color = MOI,
      group = MOI
    ),
    linewidth = 0.9
  ) +
  geom_point(
    data = efficiency_summary,
    aes(
      x = Infection.time.h,
      y = mean_EE,
      color = MOI
    ),
    size = 2.5
  ) +
  geom_errorbar(
    data = efficiency_summary %>% filter(is.finite(se_EE)),
    aes(
      x = Infection.time.h,
      ymin = mean_EE - se_EE,
      ymax = mean_EE + se_EE,
      color = MOI
    ),
    width = 0.15
  ) +
  scale_y_continuous(
    labels = scales::label_number(suffix = "%")
  ) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = "Infection time (h)",
    y = "Entrapment efficiency (%)",
    color = "MOI",
    title = ""
  ) 

# Looking at missing data
opt.df %>%
  summarise(
    missing_spores = sum(is.na(spores.ml)),
    missing_virospores = sum(is.na(virospores.ml)),
    zero_total = sum(total_spores_ml == 0, na.rm = TRUE),
    missing_time = sum(is.na(Infection.time.h)),
    valid_EE = sum(is.finite(entrapment_percent))
  )
```

4. Preliminary stats on entrapment efficiency (before realizing that the best way to analyze data is through MLE)
```{r}
opt.df$logit_entrapment <- qlogis(
  pmin(pmax(opt.df$entrapment_efficiency, 0.001), 0.999)
)

ent_model <- lm(
  logit_entrapment ~ factor(Infection.time.h) * factor(MOI),
  data = opt.df
)

anova(ent_model)

# Simplified model
ent_model2 <- lm(
  logit_entrapment ~ factor(Infection.time.h),
  data = opt.df
)

anova(ent_model2)

library(emmeans)
pairs(emmeans(ent_model2, "Infection.time.h"), adjust = "tukey")
```
2. Plot spores and virospores over time for each MOI 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Convert to long format

opt_long <- opt %>%
  pivot_longer(
    cols = c(spores.ml.corrected, virospores.ml.corrected),
    names_to = "Cell_type",
    values_to = "Count"
  ) %>%
  mutate(
    Cell_type= recode(
      Cell_type,
      spores.ml.corrected = "Spores",
      virospores.ml.corrected = "Virospores"
    )
  )

ggplot(opt_long,
       aes(x = Infection.time.h,
           y = Count,
           color = Cell_type)) +
  geom_point() +
  geom_line(aes(group = interaction(Rep, Cell_type))) +
  facet_wrap(~MOI) +
  scale_y_log10()

# PLotting with mean and SE
ggplot(opt_long,
       aes(x = Infection.time.h,
           y = Count,
           color = Cell_type)) +
  stat_summary(fun = mean,
               geom = "line",
               linewidth = 1) +
  stat_summary(fun = mean,
               geom = "point",
               size = 2.5) +
  stat_summary(fun.data = mean_se,
               geom = "errorbar",
               width = 0.4) +
  facet_wrap(~MOI) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Cell type per mL",
    color = NULL
  ) 
```
3. Preliminary stats
```{r}
opt$log_spores <- log10(opt$spores.ml.corrected + 1)
opt$log_viro <- log10(opt$virospores.ml.corrected + 1)

# Spores 
spore_model <- lm(
  log_spores ~ factor(Infection.time.h) * factor(MOI),
  data = opt
)

anova(spore_model)
# Virospores
viro_model <- lm(
  log_viro ~ factor(Infection.time.h) * factor(MOI),
  data = opt
)

anova(viro_model)


spore_model2 <- lm(
  log10(spores.ml.corrected + 1) ~ factor(Infection.time.h) + factor(MOI),
  data = opt
)

library(emmeans)
# Post hoc
emmeans(spore_model2, pairwise ~ factor(Infection.time.h), adjust = "tukey")

viro_model2 <- lm(
  log10(virospores.ml.corrected + 1) ~ factor(Infection.time.h),
  data = opt
)

anova(viro_model2)
# Post hoc pairwise comparison
pairs(emmeans(viro_model2, "Infection.time.h"), adjust = "tukey")
```
```{r}
# Additive model 
model2 <- lm(
  log_spores ~ factor(Infection.time.h) + factor(MOI),
  data = opt
)

library(emmeans)

emmeans(model2, "Infection.time.h")
pairs(emmeans(model2, "Infection.time.h"))

emmeans(model2, "MOI")
pairs(emmeans(model2, "MOI")) 
```


```{r}
ggplot(opt_long,
       aes(x = Infection.time.h,
           y = Count,
           color = Cell_type)) +
  stat_summary(fun = mean,
               geom = "line",
               linewidth = 1) +
  stat_summary(fun = mean,
               geom = "point",
               size = 2.5) +
  stat_summary(fun.data = mean_se,
               geom = "errorbar",
               width = 0.4) +
  facet_wrap(~MOI) +
  scale_color_manual(
    values = c(
      "Spores" = "blue",
      "Virospores" = "red"
    )
  ) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Cell type per mL",
    color = NULL
  )
```
Plot similarly to above
```{r}
ggplot(opt_long,
       aes(x = Infection.time.h,
           y = Count,
           color = Cell_type)) +
  stat_summary(fun = mean,
               geom = "line",
               linewidth = 1) +
  stat_summary(fun = mean,
               geom = "point",
               size = 2.5) +
  stat_summary(fun.data = mean_se,
               geom = "errorbar",
               width = 0.4) +
  facet_wrap(~MOI, nrow = 1) +
  scale_color_manual(
    values = c(
      "Spores" = "blue",
      "Virospores" = "red"
    )
  ) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Cell type per mL",
    color = NULL
  )
```
Remove NA and plot again
```{r}
ggplot(
  subset(opt_long, !is.na(Count) & !is.na(MOI)),
  aes(
    x = Infection.time.h,
    y = Count,
    color = Cell_type
  )
) +
  stat_summary(
    fun = mean,
    geom = "line",
    linewidth = 1
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    size = 2.5
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.4
  ) +
  facet_wrap(~MOI, nrow = 1) +
  scale_color_manual(
    values = c(
      "Spores" = "blue",
      "Virospores" = "red"
    )
  ) +
  scale_y_log10() +
  labs(
    x = "Time of infection (h)",
    y = "Cell type per mL",
    color = NULL
  )
```


```{r}
#Instead of calculating mean spores and se_spores on raw scale, each replicate can be converted log10 (scaling the data) THEN calculate mean and SE on the log scale, then convert back to spores per ml -- i.e., transformed each replicate to log10 averaged the transformed values, calculated error on transformed scale, then converted back to spores/mL for plotting. 
spore_summary_log <- opt %>%
  filter(spores.ml > 0) %>% # remove 0
  mutate(log_spores = log10(spores.ml)) %>% #transform the replicate
  group_by(Infection.time.h, MOI) %>% #group by 
  summarise(
    mean_log = mean(log_spores, na.rm = TRUE), #calculate average on log scale
    se_log = sd(log_spores, na.rm = TRUE) / sqrt(n()), #calculate se
    ymin = 10^(mean_log - se_log),
    ymax = 10^(mean_log + se_log),
    mean_spores = 10^mean_log,
    .groups = "drop"
  )

panel_A <- ggplot(spore_summary_log,
       aes(x = Infection.time.h,
           y = mean_spores,
           color = factor(MOI),
           group = factor(MOI))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = ymin, ymax = ymax),
    width = 0.3
  ) +
  scale_y_log10() +
  labs(
    x = "Time of Infection (h)",
    y = "Spores per mL",
    color = "MOI"
  ) 
panel_A
```

Checking data distribution 
```{r}
opt$log_spores <- log10(opt$spores.ml)
log_data <- opt %>%
  filter(spores.ml > 0) %>%
  mutate(log_spores = log10(spores.ml))

qqnorm(log_data$log_spores)
qqline(log_data$log_spores, col = "red")
```

8. Preliminary stats on uninfected spores (before I had full data, I ran this exercise)
```{r}
# in order to do anova you need to have the following: 
# indepdendent obs 
# normality 
# homogeneity of variance
# continuous dependent variable

# First determine if ANOVA is appropriate
model <- lm(log10(spores.ml + 1) ~ factor(Infection.time.h) * factor(MOI),
            data = opt)
anova(model)
par(mfrow = c(2,2))
plot(model)

# test the normality of residuals
shapiro.test(residuals(model))
```
Confirm variables and missing data
```{r}
str(opt)

colSums(is.na(opt[, c(
  "spores.ml",
  "Infection.time.h",
  "MOI"
)]))

# Remove virospore columns for now 
library(dplyr)
#opt <- opt %>%
  #select(-vPFU.count, -vPFU.dilution)

# visualize raw distributions
library(ggplot2)

ggplot(
  opt,
  aes(
    x = interaction(Infection.time.h, MOI),
    y = spores.ml
  )
) +
  geom_boxplot() +
  geom_jitter(width = 0.15) +
  scale_y_log10() +
  labs(
    x = "Infection time × MOI",
    y = "Spores per mL"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
```

Fit anova model
```{r}
opt <- opt %>%
  mutate(
    Infection.time.h = factor(Infection.time.h),
    MOI = factor(MOI)
  )

model <- lm(
  log10(spores.ml + 1) ~ Infection.time.h * MOI,
  data = opt
)

# Check residuals
par(mfrow = c(2, 2))
plot(model)
par(mfrow = c(1, 1))

# Check residual normality
qqnorm(residuals(model))
qqline(residuals(model))

shapiro.test(residuals(model)) # residuals are not normal

# Check equal variances
library(car)

leveneTest(
  log10(spores.ml + 1) ~
    interaction(Infection.time.h, MOI),
  data = opt
)
# Okay variances are equal


#Visually check the residual spread 
plot(
  fitted(model),
  residuals(model),
  xlab = "Fitted values",
  ylab = "Residuals"
)

abline(h = 0, lty = 2)

# Check for unusual obs
cooks <- cooks.distance(model)

which(cooks > 4 / nobs(model))
```
Moving forward with ANOVA
```{r}
# Fit the full model
model1 <- lm(
  log10(spores.ml + 1) ~ factor(Infection.time.h) * factor(MOI),
  data = opt
)

anova(model)
# okay so timing of infection is important and so is MOI but because the interaction between time and MOI is not significant, we can transition to an additive model 

# We can simplify the model because time and MOI were not signififcant 
model2 <- lm(
  log10(spores.ml + 1) ~ Infection.time.h + MOI, 
  data = opt
)

anova(model2)
# Post hoc tests
library(emmeans)
emm_time <- emmeans(model2, ~ Infection.time.h)
pairs(emm_time, adjust = "tukey")

# compact letter display
library(multcomp)

cld(
  emm_time,
  adjust = "tukey",
  Letters = letters
)
# Okay need to add a panel to the figure of the distribtuons of MOI-- the marginal means from emmeans for infection time time averaged across MOIOS with the compact letter display 

library(emmeans)
library(multcomp)
library(dplyr)
library(ggplot2)


# Add Tukey compact-letter groups
emm_time_letters <- cld(
  emm_time,
  adjust = "tukey",
  Letters = letters
) %>%
  as.data.frame() %>%
  mutate(
    Infection.time.h = as.numeric(as.character(Infection.time.h)),
    .group = trimws(.group)
  )

panel_B <- ggplot(
  emm_time_letters,
  aes(x = Infection.time.h, y = emmean)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower.CL, ymax = upper.CL),
    width = 0.3
  ) +
  geom_text(
    aes(y = upper.CL + 0.25, label = .group),
    size = 4
  ) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 2)
  ) +
  labs(
    x = "Infection time (h)",
    y = expression("Estimated marginal mean log"[10] *
                     "(spores mL"^{-1} * " + 1)")
  ) 

panel_B

# Because the model works on the log scale, we need to back track to spores/ML
emm_time_letters_original <- emm_time_letters %>%
  mutate(
    spores_emmean = 10^emmean - 1,
    spores_lower = 10^lower.CL - 1,
    spores_upper = 10^upper.CL - 1
  )

# Plot on tthe original scale 
panel_B <- ggplot(
  emm_time_letters_original,
  aes(x = Infection.time.h, y = spores_emmean)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = spores_lower, ymax = spores_upper),
    width = 0.3
  ) +
  geom_text(
    aes(y = spores_upper * 1.4, label = .group),
    size = 4
  ) +
  scale_x_continuous(
    breaks = seq(0, 24, by = 2)
  ) +
  scale_y_log10() +
  labs(
    x = "Infection time (h)",
    y = "Model-adjusted mean spores per mL"
  ) 

panel_B
# MOI was also significant
emm_moi <- emmeans(model, ~ MOI)
pairs(emm_moi, adjust = "tukey")

moi_results <- cld(
  emm_moi,
  adjust = "tukey",
  Letters = letters
) %>%
  as.data.frame() %>%
  mutate(
    .group = trimws(.group),
    estimated_spores = 10^emmean - 1,
    lower_spores = 10^lower.CL - 1,
    upper_spores = 10^upper.CL - 1
  )
# Plot MOIS
panel_C <- ggplot(
  moi_results,
  aes(x = MOI, y = estimated_spores)
) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = lower_spores, ymax = upper_spores),
    width = 0.15
  ) +
  geom_text(
    aes(y = upper_spores * 1.3, label = .group),
    size = 4
  ) +
  scale_y_log10() +
  labs(
    x = "MOI",
    y = "Model-adjusted mean spores per mL"
  ) 

panel_C

library(patchwork)

final_figure <- panel_A / (panel_B + panel_C) +
  plot_annotation(tag_levels = "A")

final_figure

ggsave(
  "Figure1_spores.pdf",
  width = 10,
  height = 8,
  dpi = 300
)

```
Putting the panel figures together
```{r}
library(patchwork)

combined_figure <- panel_A + panel_B +
  plot_annotation(tag_levels = "A")

combined_figure
```
Modeling time as continuous (since it shouldn't be a factor)
```{r}
model3 <- lm(log10(spores.ml + 1) ~ Infection.time.h, data = opt)
model3
anova(model3)

model4 <- lm(log10(spores.ml + 1) ~ Infection.time.h + MOI, data = opt)
anova(model4)
```