---
title: "response.surface.optimalentrap"
output: html_document
---

Response surface analysis trial-- related to https://my.che.utah.edu/~tony/course/material/ExperimentalDesign/30_meth_of_steep_asc.php 

0. Set wd and theme
```{r}
setwd("~/GitHub/VirosporeQuant/Optimal.entrap")
```

```{r}
library(paletteer)
library(ggplot2)
theme_set(theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.title.x = element_text(margin = margin(t = 15, b = 15)),
        axis.title.y = element_text(margin = margin(l = 15, r = 15)),
        axis.text = element_text(size = 13),
        axis.text.x = element_text(margin = margin(t = 5)),
        axis.text.y = element_text(margin = margin(r = 5)),
        #axis.line.x = element_line(linewidth = 1),
        #axis.line.y = element_line(linewidth = 1),
        axis.ticks.y.right = element_line(linewidth = 1),
        axis.text.y.right = element_text(size = 13, margin = margin(l = 5)),
        axis.line.y.right = element_line(linewidth = 0.1),
        axis.ticks.x.top = element_line(linewidth = 1),
        axis.text.x.top = element_text(size = 13, margin = margin(b = 5)),
        axis.line.x.top = element_line(linewidth = 0.1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_line(linewidth = 1),
        axis.ticks.y = element_line(linewidth = 1),
        axis.ticks.length = unit(.1, "in"),
        panel.border = element_rect(color = "grey20", fill = NA, linewidth = 1.5),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14),
        strip.background = element_blank()
        )) 
colors <-paletteer_d("ggthemes::Miller_Stone")
```

1. Optimal entrapment analysis for SPO1 WT phage 
```{r}
library(readxl)
opt <- read_excel("~/GitHub/VirosporeQuant/Optimal.entrap/Data/20260629_optimalentrapment_experiment_data_github.xlsx", sheet = "opt.e")
```

2. Visualize raw response surface
```{r}
library(ggplot2)
library(dplyr)

# Calculate entrapment efficiency from the data -- note that "corrected" values come from accounting for final volume of sample 
opt <- opt %>%
  mutate(
    ee = (virospores.ml.corrected / (virospores.ml.corrected + spores.ml.corrected)) * 100
  )
# Include log10 MOI
opt$log10_MOI <- log10(opt$MOI)

# Plot
ggplot(opt,
       aes(x = Infection.time.h,
           y = log10_MOI)) +
  geom_point(aes(size = ee,
                 fill = ee),
             shape = 21,
             color = "black") +
  scale_fill_viridis_c() +
  scale_y_continuous(
    breaks = c(-1, 0, 1),
    labels = c("0.1", "1", "10")
  ) +
  labs(
    x = "Time of infection (h)",
    y = "MOI",
    fill = "Entrapment efficiency",
    size = "Entrapment efficiency"
  )

```

3. Fit a first-order surface 
```{r}
# Exploratory
model_linear <- lm(
  ee ~
    Infection.time.h +
    log10_MOI,
  data = opt
)

summary(model_linear)

# Include the interaction
model_linear_int <- lm(
  ee ~
    Infection.time.h * log10_MOI,
  data = opt
)

anova(model_linear, model_linear_int)
summary(model_linear_int)
```

4. Fit a quadratic response surface model 
```{r}
# Center the factors 
opt$time_c <- opt$Infection.time.h -
  mean(opt$Infection.time.h, na.rm = TRUE)

opt$logMOI_c <- opt$log10_MOI -
  mean(opt$log10_MOI, na.rm = TRUE)

# Fit model
model_quad <- lm(
  ee ~
    time_c +
    logMOI_c +
    I(time_c^2) +
    I(logMOI_c^2) +
    time_c:logMOI_c,
  data = opt
)

summary(model_quad)
anova(model_linear_int, model_quad)

# More appropriate model based on modeling the counts 
model_entrapment <- glm(
  cbind(virospores.ml, spores.ml) ~
    time_c +
    logMOI_c +
    I(time_c^2) +
    I(logMOI_c^2) +
    time_c:logMOI_c,
  family = quasibinomial,
  data = opt
)

summary(model_entrapment)
```
5. Predict the optimum over the tested region
```{r}
prediction_grid <- expand.grid(
  Infection.time.h = seq(
    min(opt$Infection.time.h),
    max(opt$Infection.time.h),
    length.out = 200
  ),
  MOI = 10^seq(
    log10(min(opt$MOI, na.rm = TRUE)),
    log10(max(opt$MOI, na.rm = TRUE)),
    length.out = 200
  )
)

prediction_grid$time_c <-
  prediction_grid$Infection.time.h -
  mean(opt$Infection.time.h, na.rm = TRUE)

prediction_grid$logMOI_c <-
  log10(prediction_grid$MOI) -
  mean(opt$log10_MOI, na.rm = TRUE)

prediction_grid$predicted_entrapment <- predict(
  model_entrapment,
  newdata = prediction_grid,
  type = "response"
)
```

6. Find the predicted maximum
```{r}
predicted_optimum <- prediction_grid[
  which.max(prediction_grid$predicted_entrapment),
]

predicted_optimum
```

7. Display the surface
```{r}
ggplot(
  prediction_grid,
  aes(
    x = Infection.time.h,
    y = log10(MOI),
    z = predicted_entrapment
  )
) +
  geom_contour_filled() +
  geom_point(
    data = opt,
    aes(
      x = Infection.time.h,
      y = log10(MOI)
    ),
    inherit.aes = FALSE
  ) +
  scale_y_continuous(
    breaks = c(-1, 0, 1),
    labels = c("0.1", "1", "10")
  ) +
  labs(
    x = "Time of infection (h)",
    y = "MOI",
    fill = "Predicted\nentrapment"
  )
```

# Okay probably should not have been modeling with entrapment efficiency but rather virospore production (?)

```{r}
opt$logMOI <- log10(opt$MOI)

# Center predictors
opt$time_c <- opt$Infection.time.h -
  mean(opt$Infection.time.h, na.rm = TRUE)

opt$logMOI_c <- opt$logMOI -
  mean(opt$logMOI, na.rm = TRUE)

rsm_model <- lm(
  log10(virospores.ml.corrected + 1) ~
    time_c +
    I(time_c^2) +
    logMOI_c +
    I(logMOI_c^2) +
    time_c:logMOI_c,
  data = opt
)

summary(rsm_model)
anova(rsm_model)
```

```{r}
m1 <- lm(
  log10(virospores.ml.corrected + 1) ~ time_c + logMOI_c,
  data = opt
)

m2 <- lm(
  log10(virospores.ml.corrected + 1) ~ time_c * logMOI_c,
  data = opt
)

m3 <- lm(
  log10(virospores.ml.corrected + 1) ~
    time_c +
    I(time_c^2) +
    logMOI_c +
    I(logMOI_c^2) +
    time_c:logMOI_c,
  data = opt
)

anova(m1, m2, m3)
AIC(m1, m2, m3)
```

```{r}
par(mfrow = c(2, 2))
plot(m3)
par(mfrow = c(1, 1))
```


```{r}
library(ggplot2)
# Prepare variables
opt$logMOI <- log10(opt$MOI)

opt$time_c <- opt$Infection.time.h -
  mean(opt$Infection.time.h, na.rm = TRUE)

opt$logMOI_c <- opt$logMOI -
  mean(opt$logMOI, na.rm = TRUE)

# Quadratic response-surface model
viro_surface <- lm(
  log10(virospores.ml + 1) ~
    time_c +
    I(time_c^2) +
    logMOI_c +
    I(logMOI_c^2) +
    time_c:logMOI_c,
  data = opt
)

summary(viro_surface)
anova(viro_surface)
```
```{r}
par(mfrow = c(2, 2))
plot(viro_surface)
par(mfrow = c(1, 1))
```


```{r}
viro_linear <- lm(
  log10(virospores.ml + 1) ~
    time_c + logMOI_c,
  data = opt
)

viro_interaction <- lm(
  log10(virospores.ml + 1) ~
    time_c * logMOI_c,
  data = opt
)

anova(
  viro_linear,
  viro_interaction,
  viro_surface
)

AIC(
  viro_linear,
  viro_interaction,
  viro_surface
)
```



```{r}
prediction_grid <- expand.grid(
  Infection.time.h = seq(
    min(opt$Infection.time.h, na.rm = TRUE),
    max(opt$Infection.time.h, na.rm = TRUE),
    length.out = 200
  ),
  logMOI = seq(
    min(opt$logMOI, na.rm = TRUE),
    max(opt$logMOI, na.rm = TRUE),
    length.out = 200
  )
)

prediction_grid$time_c <-
  prediction_grid$Infection.time.h -
  mean(opt$Infection.time.h, na.rm = TRUE)

prediction_grid$logMOI_c <-
  prediction_grid$logMOI -
  mean(opt$logMOI, na.rm = TRUE)

# Predictions are initially on the log10 scale
prediction_grid$predicted_log_virospores <- predict(
  viro_surface,
  newdata = prediction_grid
)

# Back-transform to virospores/mL
prediction_grid$predicted_virospores_ml <-
  10^prediction_grid$predicted_log_virospores - 1
```

```{r}
ggplot(
  prediction_grid,
  aes(
    x = Infection.time.h,
    y = logMOI,
    z = predicted_log_virospores
  )
) +
  geom_contour_filled() +
  geom_point(
    data = opt,
    aes(
      x = Infection.time.h,
      y = logMOI
    ),
    inherit.aes = FALSE,
    size = 2
  ) +
  scale_y_continuous(
    breaks = c(-1, 0, 1),
    labels = c("0.1", "1", "10")
  ) +
  labs(
    x = "Time of infection (h)",
    y = "MOI",
    fill = expression(
      "Predicted log"[10] * "(virospores/mL + 1)"
    )
  ) 
```
```{r}
predicted_optimum <- prediction_grid[
  which.max(prediction_grid$predicted_virospores_ml),
]

predicted_optimum


predicted_optimum$MOI <- 10^predicted_optimum$logMOI

predicted_optimum[
  c(
    "Infection.time.h",
    "MOI",
    "predicted_virospores_ml"
  )
]
```

```{r}

library(plotly)

# Values used to create the prediction grid
time_vals <- sort(unique(prediction_grid$Infection.time.h))
moi_vals  <- sort(unique(prediction_grid$logMOI))

# Convert predicted values into a matrix
z_matrix <- matrix(
  prediction_grid$predicted_log_virospores,
  nrow = length(moi_vals),
  ncol = length(time_vals),
  byrow = TRUE
)

# Make 3D surface
plot_ly(
  x = time_vals,
  y = 10^moi_vals,
  z = z_matrix,
  type = "surface"
) %>%
  layout(
    scene = list(
      xaxis = list(
        title = "Infection time (h)"
      ),
      yaxis = list(
        title = "MOI",
        type = "log"
      ),
      zaxis = list(
        title = "Predicted log10(virospores/mL + 1)"
      )
    )
  )
plot_ly(
  x = time_vals,
  y = moi_vals,
  z = z_matrix,
  type = "surface"
) %>%
  layout(
    scene = list(
      xaxis = list(
        title = "Infection time (h)"
      ),
      yaxis = list(
        title = "MOI",
        tickvals = c(-1, 0, 1),
        ticktext = c("0.1", "1", "10")
      ),
      zaxis = list(
        title = "Predicted log10(virospores/mL + 1)"
      )
    )
  )

```
```

