---
title: "Virus entrapment"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Virus entrapment dynamics

## Clear workspace and set directory

```{r setup}
rm(list = ls())
getwd()
setwd("~/Desktop/Virospore/")
```

## Load packages and functions

```{r}
require("png")
require("dplyr")
require("grid")
require("plyr")
require("tidyverse")
require("bbmle")
```

## Load data

```{r}
viro <- read.csv("virospore.csv", header = TRUE, sep = ",")

# Subset data for MOI = 0.1
viro.sub <- viro %>%
  filter(MOI == 0.1, !is.na(virospores.ml)) %>%
  mutate(
    time = Infection.time.h,
    y = log10(virospores.ml)
  )

# Check the data used in the analysis
nrow(viro.sub)
summary(viro.sub)

# Observed range of infection time
range(viro.sub$time)
```

## Fit equations with maximum likelihood

### Model 1: Flat baseline + symmetric Gaussian window

```{r}
# Starting values
B0 = min(viro.sub$y)
A = max(viro.sub$y) - min(viro.sub$y)
MU = viro.sub$time[which.max(viro.sub$y)]
SIGMA = 4
Z = sd(viro.sub$y)

fit1 <- mle2(y ~
    dnorm(mean = b0 +
          a * exp(-(time - mu)^2 / (2 * sigma^2)),
          sd = z),
    start = list(b0 = B0, a = A, mu = MU,
                 sigma = SIGMA, z = Z),
    method = "L-BFGS-B",
    lower = c(b0 = -Inf,
              a = 0,
              mu = min(viro.sub$time),
              sigma = 0.001,
              z = 0.001),
    upper = c(b0 = Inf,
              a = Inf,
              mu = max(viro.sub$time),
              sigma = Inf,
              z = Inf),
    data = viro.sub)

summary(fit1)
```

### Model 2: Linear baseline + symmetric Gaussian window

```{r}
# Starting values
B0 = min(viro.sub$y)
B1 = -0.05
A = max(viro.sub$y) - min(viro.sub$y)
MU = viro.sub$time[which.max(viro.sub$y)]
SIGMA = 4
Z = sd(viro.sub$y)

fit2 <- mle2(y ~
    dnorm(mean = b0 + b1 * time +
          a * exp(-(time - mu)^2 / (2 * sigma^2)),
          sd = z),
    start = list(b0 = B0, b1 = B1, a = A, mu = MU,
                 sigma = SIGMA, z = Z),
    method = "L-BFGS-B",
    lower = c(b0 = -Inf,
              b1 = -Inf,
              a = 0,
              mu = min(viro.sub$time),
              sigma = 0.001,
              z = 0.001),
    upper = c(b0 = Inf,
              b1 = Inf,
              a = Inf,
              mu = max(viro.sub$time),
              sigma = Inf,
              z = Inf),
    data = viro.sub)

summary(fit2)
```

### Model 3: Flat baseline + asymmetric Gaussian window

```{r}
# Starting values
B0 = min(viro.sub$y)
A = max(viro.sub$y) - min(viro.sub$y)
MU = viro.sub$time[which.max(viro.sub$y)]
SIGMAL = 4
SIGMAR = 4
Z = sd(viro.sub$y)

fit3 <- mle2(y ~
    dnorm(mean = b0 +
          a * ifelse(time <= mu,
              exp(-(time - mu)^2 / (2 * sigmaL^2)),
              exp(-(time - mu)^2 / (2 * sigmaR^2))),
          sd = z),
    start = list(b0 = B0, a = A, mu = MU,
                 sigmaL = SIGMAL, sigmaR = SIGMAR, z = Z),
    method = "L-BFGS-B",
    lower = c(b0 = -Inf,
              a = 0,
              mu = min(viro.sub$time),
              sigmaL = 0.001,
              sigmaR = 0.001,
              z = 0.001),
    upper = c(b0 = Inf,
              a = Inf,
              mu = max(viro.sub$time),
              sigmaL = Inf,
              sigmaR = Inf,
              z = Inf),
    data = viro.sub)

summary(fit3)
```

## Compare models

```{r}
# Compare models using AIC
AIC(fit1, fit2, fit3)

# Calculate delta AIC
aic <- AIC(fit1, fit2, fit3)

aic$deltaAIC <- aic$AIC - min(aic$AIC)

# Calculate AIC weights
aic$weight <- exp(-0.5 * aic$deltaAIC) /
              sum(exp(-0.5 * aic$deltaAIC))

aic
```


## Inferring entrapment window parameters from the estimates in fit 3

The asymmetric Gaussian model (fit 3) provides estimates of the timing and breadth of
the entrapment window. The peak occurs at `mu`. The breadth of the window is
described by the half-width at half-maximum (HWHM) on the early and late sides,
and by their sum, the full width at half-maximum (FWHM).

```{r}
# Extract coefficients from the best model
p <- coef(fit3)

# Extract parameters
mu <- p["mu"]
sigmaL <- p["sigmaL"]
sigmaR <- p["sigmaR"]

# Conversion from sigma to half-width at half-maximum
hwhm <- sqrt(2 * log(2))

# Calculate window parameters
HWHM_L <- hwhm * sigmaL
HWHM_R <- hwhm * sigmaR
FWHM <- HWHM_L + HWHM_R

# Asymmetry index
asymmetry <- (sigmaR - sigmaL) / (sigmaR + sigmaL)

# Ratio of right to left width
width_ratio <- sigmaR / sigmaL

# Half-maximum times
half_left <- mu - HWHM_L
half_right <- mu + HWHM_R

# Create table of window parameters
window <- data.frame(
  Parameter = c(
    "Peak time (h)",
    "Left HWHM (h)",
    "Right HWHM (h)",
    "FWHM (h)",
    "Left half-maximum time (h)",
    "Right half-maximum time (h)",
    "Asymmetry index",
    "Right / left width"
  ),
  Estimate = c(
    mu,
    HWHM_L,
    HWHM_R,
    FWHM,
    half_left,
    half_right,
    asymmetry,
    width_ratio
  )
)

window
```

## Plot data and fit3 predicted values

```{r}
png(filename="~/Desktop/Virospore/fit3_moi0.1.png",
    width = 1200, height = 1200, res = 96*2)

plot.new()
par(mar = c(7, 7, 5, 7))

plot(viro.sub$time, viro.sub$virospores.ml,
     xlim = c(-1, 26),
     ylim = c(1e3, 1e12),
     type = "p",
     log = "y",
     pch = 22,
     bg = "red",
     col = "red",
     cex = 2,
     ylab = "",
     xlab = "",
     las = 1,
     lwd = 2,
     yaxt = "n",
     xaxt = "n")

box(lwd = 2)

# Add y-axis ticks and labels
axis(side = 2,
     lwd.ticks = 2,
     las = 1,
     cex.axis = 1.25,
     at = 10^(3:12),
     labels = expression(
       10^3, 10^4, 10^5, 10^6, 10^7,
       10^8, 10^9, 10^10, 10^11, 10^12
     ))

axis(side = 4,
     labels = FALSE,
     lwd.ticks = 2,
     at = 10^(3:12))

# Add x-axis ticks and labels
axis(side = 1,
     lwd.ticks = 2,
     cex.axis = 1.25,
     las = 1,
     mgp = c(3, 1, 0),
     labels = c(0, 5, 10, 15, 20, 25),
     at = c(0, 5, 10, 15, 20, 25))

axis(side = 3,
     labels = FALSE,
     lwd.ticks = 2,
     at = c(0, 5, 10, 15, 20, 25))

# Axis titles
mtext("Infection time (h)",
      side = 1,
      outer = TRUE,
      cex = 1.5,
      line = -4,
      adj = 0.5)

mtext(expression(paste("Virospores (mL"^{-1}, ")")),
      side = 2,
      outer = TRUE,
      cex = 1.5,
      line = -3.5,
      adj = 0.55)

# Extract fit3 coefficients
p <- coef(fit3)

# Infection times for predicted curve
x.pred <- seq(0, 24, length.out = 1000)

# Predicted log10 virospore concentration
y.pred.log <- p["b0"] +
  p["a"] *
  ifelse(
    x.pred <= p["mu"],
    exp(-(x.pred - p["mu"])^2 /
          (2 * p["sigmaL"]^2)),
    exp(-(x.pred - p["mu"])^2 /
          (2 * p["sigmaR"]^2))
  )

# Convert predictions to original virospore scale
y.pred <- 10^y.pred.log

# Add fitted curve
lines(x.pred, y.pred,
      col = "red",
      lty = 1,
      lwd = 3)

# Close plot device
dev.off()
graphics.off()

# Show plot
img <- readPNG("~/Desktop/Virospore/fit3_moi0.1.png")
grid.raster(img)
```


